

---
title: "Sparse Graphical Models with Censored or Missing Data using GHScenmis"
author: "The Tien Mai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse Graphical Models with Censored or Missing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ghscenmis)
```
---
```
# 1 Introduction

The **GHScenmis** package provides methods for estimating **sparse precision matrices**, also known as Gaussian graphical models, when data are affected by **censoring** or **missingness**. Conventional sparse estimators assume complete data, but many real-world datasets contain observations that fall below/above detection limits or are missing at random.

This vignette demonstrates how to use the package for:

1. Censored multivariate Gaussian data
2. Data with missing values

Both procedures rely on the **Graphical Horseshoe (GHS)** prior with an efficient Gibbs sampler.

# 2 Installation

The development version can be installed from GitHub:

```r
# install.packages("devtools")
devtools::install_github("tienmt/ghscenmis")
library(ghscenmis)
```

# 3 Example: Censored Data

This section illustrates fitting a sparse precision matrix when the data are **left-censored**. We generate synthetic multivariate normal data, apply censoring, and fit the censored GHS model.

## 3.1 Data generation

```{r example-censored-data}
set.seed(123)
p <- 10
n <- 200

# True covariance with a 3-variable block
Sigma_true <- diag(p)
Sigma_true[1, 2] <- Sigma_true[2, 1] <-
  Sigma_true[1, 3] <- Sigma_true[3, 1] <-
  Sigma_true[2, 3] <- Sigma_true[3, 2] <- 0.5

Omega_true <- solve(Sigma_true)

# Generate data
Z <- MASS::mvrnorm(n, rep(0, p), Sigma_true)

# Apply left-censoring at variable-specific thresholds
cvec <- rep(c(-0.5, 0.5), length.out = p)
Y <- pmax(Z, matrix(cvec, n, p, byrow = TRUE))
```

## 3.2 Fitting the censored GHS model

```{r fit-censored}
fit <- censored_GHS(
  Ytilde = Y,
  cvec = cvec,
  n_iter = 3000,
  burn = 500,
  verbose = FALSE
)

# Posterior mean precision matrix
Omega_hat <- fit$Omega_mean

# Edge selection based on posterior credible intervals
edges <- get_edge_selection_ci(fit$Omega_draws)$E_est
```

## 3.3 Evaluating performance

```{r evaluate-censored}
# True adjacency
E_true <- 1 * (abs(Omega_true) > 1e-6)
diag(E_true) <- 0

metrics <- get_edge_metrics(E_true, edges)
metrics
```

The results include common edge selection metrics such as TPR, FPR, Precision, and F1.
Users may also examine posterior draws or credible intervals for deeper analysis.

# 4 Example: Missing Data

The package also provides a Gibbs sampler for datasets containing **missing values** (coded as `NA`). Missing entries are imputed internally.

## 4.1 Generating incomplete data

```{r data-missing}
set.seed(123)
Y <- matrix(rnorm(100 * 5), 100, 5)
Y[sample(length(Y), 10)] <- NA  # introduce missing values
```

## 4.2 Fitting the missing-data GHS model

```{r fit-missing}
fit_mis <- ghs_missing_gibbs(Y = Y)

Omega_mis <- fit_mis$Omega_mean
Omega_mis
```

Posterior samples of the precision matrix are stored in `fit_mis$Omega_draws`, and edge selection can be performed similarly using:

```r
get_edge_selection_ci(fit_mis$Omega_draws)
```

# 5 Key Functions

The following functions provide the core functionality:

| Function                  | Purpose                                                          |
| ------------------------- | ---------------------------------------------------------------- |
| `censored_GHS()`          | Estimate sparse precision matrices with censored data.           |
| `ghs_missing_gibbs()`     | Gibbs sampler for sparse estimation with missing values.         |
| `get_edge_selection_ci()` | Edge selection using posterior credible intervals.               |
| `get_edge_metrics()`      | Standard edge-selection metrics such as TPR, FPR, Precision, F1. |

# 6 References

Li, Y., Craig, B. A., & Bhadra, A. (2019).
*The graphical horseshoe estimator for inverse covariance matrices.*
Journal of Computational and Graphical Statistics, 28(3), 747–757.


